Vertical and oblique transmission with fluctuating transmission

### 2018-03-15 ms/fluctuating_oblique.tex
merged notebooks; fixed UL comments

We now consider the case that both transmission and selection fluctuate randomly.
Rewrite eq.~28 from \citet{Ram2018} so that $\rho$ is also a random variable
\begin{equation}
x_t = x_t \frac{1 + \rho_t s_t + x_t (1 - \rho_t) s_t}{1 + x_t s_t},
\end{equation}
where $s_t$ are independent and identically distributed and also $\rho_t$ are independent and identically distributed ($t=0,1,2,\ldots$), $P(-1+C<s_t<D)=1$ for some positive $C$ and $D$, and $0<\rho_t<1$.
Therefore, $z_t = \rho_t s_t$  are independent and identically distributed and $P(-1+C < z_t < D$ and from Result~6 and 7 of \citet{Ram2018} we have:
\begin{itemize}
\item Suppose $E[log(1+\rho_t s_t)]>0$. Then $x^*=0$ is not stochastically locally stable and in fact $P(lim_{t \to \infty} x_t=0) = 0$, i.e., fixation of $B$ almost surely does not occur.
\item Suppose $E[log(1+\rho_t s_t)]<0$. Then $x^*=0$ is stochastically locally stable and 
\item Similarly, $E[log(1-\rho_t s_t/(1+s_t)]<0$ leads to stochastic local stability of $x^*=1$, and $E[log(1-\rho_t s_t/(1+s_t))]>0$ assures that fixation of $A$ almost surely does not occur.
\item In particular, if $E[\rho_t s_t] = cov(\rho_t, s_t) + E[\rho_t] E[s_t] \le 0$ then $x^*=0$ is stochastically locally stable, and similarly if $E[-\rho_t s_t/(1+s_t)] = cov(\rho_t, -s_t/(1+s_t)) - E[\rho_t] E[s_t/(1+s_t)] \le 0$ then $x^*=1$ is stochastically locally stable.
\end{itemize}

We will now present a number of examples.
First, if $s_t \sim U(-1, 1)$ and $\rho_t \sim U(0,1)$ independently (in particular, $cov(\rho_t, s_t)=0$), then both $E[log(1+\rho_t s_t)]$ and $E[log(1-\rho_t s_t/(1+s_t))]$ are negative, and therefore both fixations ($A$ and $B$) are stochastically locally stable. 
It is important to note that without fluctuations in $\rho$ (i.e. fixed $\rho$) it is not possible to have both fixations stable~\citep{Ram2018}.

Second, if $s_t$ is symmetric around zero ($E[s_t]=E[s_t/(1+s_t)]=0$), and $\rho_t$ and $s_t$ are positively (negatively) correlated, such that vertical transmission is more likely when $A$ ($B$) is favored, then $E[-\rho_t s_t/(1+s_t)]<0$ ($E[-\rho_t s_t]<0$) and therefore fixation of $A$ ($B$) is stochastically locally stable.
For example, let $s_t \sim U(-1,1)$ and $\rho_t \sim \beta(1+s, 1)$, such that the covariance of $s_t$ and $\rho_t$ is positive ($cov(s_t, \rho_t) \approx 0.22$), that is, vertical transmission is more likely when $A$ if favored (when $s_t>0$) and oblique transmission is more  likely when $B$ is favored (when $s_t<0$), then ($Beta$ being the beta function), 
\begin{equation}
E[log(1+\rho_t s_t)] = \frac{1}{2} \int_{-1}^{1} {\int_0^1 {\frac{\rho^s \log{(1+\rho s)}}{Beta(1+s, 1)}  \; d\rho}\; ds} \approx 0.2525,
\end{equation}
and $B$ almost surely doesn't fix, and
\begin{equation}
E[log(1-\rho_t s_t / (1+s_t)] = \frac{1}{2} \int_{-1}^{1} {\int_0^1 {\frac{(1-\rho)^{-s/(1+s)} \log{(1-\rho s / (1+s))}}{Beta(1, 1-s/(1+s)}  \; d\rho}\; ds} \approx -0.076,
\end{equation}
so fixation of $A$ is stochastically locally stable.
The opposite occurs if $\rho_t \sim \beta(1+s, 1)$ and therefore the covariance of $s_t$ and $\rho_t$ is negative ($cov(s_t, \rho_t) \approx -0.2$); in that case, fixation of $B$ is stochastically locally stable and $A$ almost surely doesn't fix.

Third, it is also possible that both fixations are stochastically locally unstable, but, similar to the case of periodic fluctuations, this can only occur if the fluctuations in $\rho_t$ are small.
For example, \autoref{fig:rho1_rho2_stoch_p} shows the expected outcome when $s_t=s$ and $\rho_t=\rho_1$ with probability $p=0.505$ and $s_t=-s$ and $\rho_t=\rho_2$ with probability $1-p=0.495$. The blue and red areas denote expected fixation of $A$ and $B$, respectively (i.e. stochastic local stability) and the white area shows expected polymorphism (both fixations are not stochastically locally stable).

>>>>

We now consider the case that both transmission and selection fluctuate randomly.
Rewrite eq.~28 from \citet{Ram2018} so that $\rho$ is also a random variable
\begin{equation}
x_t = x_t \frac{1 + \rho_t s_t + x_t (1 - \rho_t) s_t}{1 + x_t s_t},
\end{equation}
where $s_t$ are independent and identically distributed and also $\rho_t$ are independent and identically distributed ($t=0,1,2,\ldots$), $P(-1+C<s_t<D)=1$ for some positive $C$ and $D$, and $0<\rho_t<1$.
Therefore, $z_t = \rho_t s_t$  are independent and identically distributed and $P(-1+C < z_t < D$ and from Result~6 and 7 of \citet{Ram2018} we have:
\begin{itemize}
\item Suppose $E[log(1+\rho_t s_t)]>0$. Then $x^*=0$ is not stochastically locally stable and in fact $P(lim_{t \to \infty} x_t=0) = 0$, i.e., fixation of $B$ almost surely does not occur.
\item Suppose $E[log(1+\rho_t s_t)]<0$. Then $x^*=0$ is stochastically locally stable and 
\item Similarly, $E[log(1-\rho_t s_t/(1+s_t)]<0$ leads to stochastic local stability of $x^*=1$, and $E[log(1-\rho_t s_t/(1+s_t))]>0$ assures that fixation of $A$ almost surely does not occur.
\item In particular, if $E[\rho_t s_t] = cov(\rho_t, s_t) + E[\rho_t] E[s_t] \le 0$ then $x^*=0$ is stochastically locally stable, and similarly if $E[-\rho_t s_t/(1+s_t)] = cov(\rho_t, -s_t/(1+s_t)) - E[\rho_t] E[s_t/(1+s_t)] \le 0$ then $x^*=1$ is stochastically locally stable.
\item Note that it is not possible that both $E[log(1+\rho_t s_t)]$ and $E[log(1-\rho_t s_t/(1+s_t)]$ are negative, as their sum is positive
\begin{multline}
E[log(1+\rho_t s_t)] + E[log(1-\rho_t s_t/(1+s_t)] = \\
E[log(1+\rho_t s_t) + log(1-\rho_t s_t/(1+s_t)]= \\
E[log\big((1+\rho_t s_t)(1-\rho_t s_t/(1+s_t)\big)]= \\
E[log\big( 1+\rho_t(1-\rho_t)(s_t^2)/(1+s_t) \big)] > 0,
\end{multline}
and therefore it is not possible that both fixations are stochastically locally stable.
\end{itemize}

We will now present a number of examples.
First, if $s_t$ and $\rho_t$ are independent ($cov(s_t, \rho_t)$) and $s_t$ is symmetric around zero, then $E[\rho_t s_t]=0$ and $E[-\rho_t s_t/(1+s_t)]>0$ (because $E[s_t/(1+s_t)] < E[s_t]$).
Therefore, fixation of $B$ is stochastically locally stable and fixation of $A$ almost surely doesn't occur.
For example, let $s_t \sim U(-1, 1)$ and $\rho_t \sim U(0,1)$ independently (in particular, $cov(\rho_t, s_t)=0$), then $E[log(1+\rho_t s_t)]\approx -0.07315$ and $E[log(1-\rho_t s_t/(1+s_t))]\approx 0.2337$.
However, note that symmetry of $s_t$ around zero provides an advantage to phenotype $B$ -- using Jensen's inequality, $E[w_A/w_B] = E[1+s_t] = 1 \le E[1/(1+s_t)] = E[w_B/w_A]$.
Therefore, if we take $w_{A,t}, w_{B,t} \sim U(0,1)$ independently, and $s_t=(w_A-w_B)/w_B$, then neither $A$ nor $B$ has an advantage, on average (i.e. $E[w_A/w_B]=E[w_B/w_A]$, see \autoref{fig:beta}A), and both $E[\rho_t s_t]$ and $E[-\rho_t s_t/(1+s_t)]$ are positive, so that both fixations are not stochastically locally stable.

If instead we do not assume independence of $s_t$ and $\rho_t$, we can get fixation. Let $\rho_t \sim \beta(1+s, 1)$, where $s_t=(w_A-w_B)/w_B$ and $w_A, w_B \sim U(0,1)$, such that the covariance of $s_t$ and $\rho_t$ is positive ($cov(s_t, \rho_t) \approx 4$), that is, vertical transmission is more likely when $A$ if favored (when $s_t>0$) and oblique transmission is more  likely when $B$ is favored (when $s_t<0$; \autoref{fig:beta}).
Then $E[log(1+\rho_t s_t)] >0$ and $B$ almost surely doesn't fix, and $E[log(1-\rho_t s_t / (1+s_t)] < 0$, so fixation of $A$ is stochastically locally stable.
The opposite occurs if $\rho_t \sim \beta(1, 1+s)$ and the covariance of $s_t$ and $\rho_t$ is negative ($cov(s_t, \rho_t) \approx -4$). In that case, fixation of $B$ is stochastically locally stable and $A$ almost surely doesn't fix.

\begin{figure*}[hbt]
\centering
\includegraphics[width=\linewidth]{../figures/{beta}.png}
\caption{
\textbf{Covariance of selection and transmission.}
\textbf{(A)} Histogram of $w_A/w_B$ where $w_A$ and $w_B$ are identically and independently distributed uniform random variables $U(0,1)$.
\textbf{(B)} Histogram of $s_t = (w_A-w_B)/w_B$.
\textbf{(C)} Histogram of $\rho_t \sim Beta(1+s_t, 1)$.
\textbf{(C)} The joint distribution of $\rho_t$ and $s_t$ demonstrates a positive correlation $cov(s_t, \rho_t)>0$.
}
\label{fig:beta}
\end{figure*}

Third, it is also possible that both fixations are not stochastically locally stable even if $s_t$ and $\rho_t$ covary, but, similar to the case of periodic fluctuations, this can only occur if the fluctuations in $\rho_t$ are small.
For example, \autoref{fig:rho1_rho2_stoch_p} shows the expected outcome when $s_t=s$ and $\rho_t=\rho_1$ with probability $p=0.505$ and $s_t=-s$ and $\rho_t=\rho_2$ with probability $1-p=0.495$.
The blue and red areas denote expected fixation of $A$ and $B$, respectively (i.e. stochastic local stability) and the white area shows expected polymorphism (both fixations are not stochastically locally stable).
